A Kolmogorov zero to one law for Markov Categories

### 2019-10-11 categorical-zero-one.tex
added sketch of hewitt-savage argument

>>>>

We can also prove a version of the Hewitt-Savage theorem
\begin{definition}
    Let $\alpha: J \to J$ be any map.
    Then it induces a map 
    \[]\hat{\alpha}: \bigotimes_{i\in J}X_i \to \bigotimes_{i \in J}X_{\alpha(i)},\]
    which we may suggestively write as $(x_i)_{i\in J} \mapsto (x_{\alpha(i)})_{i\in J}$.

    It is clear that this map is always deterministic. Moreover, if $\alpha$ is a bijection, this map is an isomorphism.
\end{definition}

\begin{theorem}[Hewitt-Savage for Markov categories]
    Let $\cC$ be a causal Markov category with infinite tensor products. Let $A \to \cC$ be a map in $\cC$, $J$ an infinite set, and let $f: \bigotimes_{i \in J}X \to T$ be a deterministic map.
    Suppose for each finite permutation $\sigma: J \to J$, we have $f \circ \hat{\sigma} = f$ - in other words, $f$ is independent of finite permutations of the inputs.
    Then the composite $A \to \tensor_J A \to \tensor_J X \to T$ is deterministic.
\end{theorem}
Again, we factor part of the proof into a lemma.
\begin{lemma}
    Let $\cC$ be a causal markov category, and let $f,g: B \to X$ be maps, $f$ deterministic.
    Let $p: A \to B$ be any map.
    Suppose the composites 
    \[A \to B \to B \tensor B \labelto{f \tensor g} X \tensor X\]
    and
    \[A \to B \to B \tensor B \labelto{f \tensor f} X \tensor X\]
    agree. Then $f = g$ $p$-almost everywhere.
\end{lemma}
\begin{proof}[Sketch of sketch of proof]
    Consider the map
    $A \labelto{p} B \to B \tensor B \labelto{f \tensor g} X \tensor X \labelto{copy_{X \tensor X}} X \tensor X \tensor X \tensor X$.
    Whether we marginalize on the first or second factor of $X$, we get equal morphisms, because we may replace the first part of this with
    $A \to B \labelto{f} B \to X \tensor X \tensor X \tensor X$ by using the assumptions and determinism.

    Then applying causality, with $f = p$, $g= (f \tensor g) \circ copy_B$, $h_1 = (del_X \tensor 1_X)$, $h_2  = (1_X \tensor del_X)$.
    Marginalizing the resulting (equal) diagrams, one of them is 
    \[A \to B \to B \tensor B \labelto{f, 1_B} B \tensor B,\]
    one of them is 
    \[A \to B \to B \tensor B \labelto{g,1_B} B \tensor B,\]
    and this is precisely what must hold for them to agree $p$-a.e.
\end{proof}

\begin{proof}[Proof of theorem]
    Consider the map

    \[e: A \to \tensor_J A \labelto{\tensor p} (\tensor_J X) \to (\tensor_J X) \tensor (\tensor_J X) \to T \tensor (\tensor_J X)\]

    Denote by $e_\sigma$ the composite of $e$ with $1_T \tensor \hat{\sigma}$
    Note that if $\sigma$ is a finite permutation, $e_\sigma = e$. To see this, note that we may add a $\hat{\sigma}$ before the $f$ as well, by assumption.
    Now we can use the determinism of the $\hat{\sigma}$s, and the clear fact that 
    \[A \to \tensor A \to \tensor B \labelto{\hat{\sigma}} \tensor B = A \to \tensor A \to \tensor B\]

    Now let $\sigma$ instead be an injection. Then for each finite subset $F \subset J$, we can find a finite permutation of $J$, $\sigma_0$, such that $\sigma|_F = \sigma_0|_F$.
    Then on this marginal, $e_\sigma = e_{\sigma_0} = e$, hence $e_\sigma = e$ in general.

    Let $J = J_0 \sqcup J_1$ be a decomposition of $J$ into two disjoint subsets with the cardinality of $J$ (here we use $J$ infinite).
    Let $\tau_0, \tau_1: J \to J$ be injections with image $J_0,J_1$ respectively.

    Now consider the map $a_\sigma: A \to T \tensor T$ given by composing $e_\sigma$ with $t$.
    By the above, they all agree.
    By the lemma, we see that the maps $\tensor_J X \labelto{\hat{\sigma}} \tensor_J X \to T$ are all $\tensor_J p$ a.e. equal.
    Applying this to $\tau_0,\tau_1$, we get that the map $A \to T$ must be deterministic.
\end{proof}
In the last step, the idea is that precomposing with $\hat{\tau_0}$ and $\hat{\tau_1}$ is essentially picking out two disjoint subsets and using them to determine $T$.
Since all the variables are independent, these two are clearly independent, but also equal (by the preceding argument). Hence they are deterministic, but they also agree with the original map.
